Extended Data

### 2021-04-30 doc/manuscript/Dow_Extended_Data.Rmd
version sent to coauthors for initial review

pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Relationship between growth parameters and mean maximum temperature within the critical temperature window (CTW).** CTW was defined using *climwin*. For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!

>>>>

pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Relationship between growth parameters and mean maximum temperature within the critical temperature window (CTW).** CTW was defined using *climwin*. For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

### 2021-04-30 doc/manuscript/Dow_manuscript_Nature.Rmd
version sent to coauthors for initial review

**Title:** Warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity

@@

DOY_timing_caption <- "**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product(**YEARS**), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. (**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50). )" #needs attention!!!

DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25, 50 and 75% of annual growth is acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) decline significantly with mean $T_{max}$ during the critical temperature window (CTW). For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands. (**DOYrmax to be dropped. NOTE THAT THIS PARTIALLY REPLICATES EXTENDED DATA FIG 4**)" #needs attention!!!

@@

@@

***(CONDENSE THE FOLLOWING 2 PARAGRAPHS INTO 1, AND REMOVE REPETITION WITH ABSTRACT:)***

**In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].**
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009].
The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].

The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019].
*This expectation is represented in models [@zani_increased_2020].*
However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].

**We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures.** 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview]. *(We need to make sure we're not missing anything.)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
Thus, tree-ring records do not necessarily corroborate findings that warming spring temperatures lengthen the growing season and increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

**Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.** 
First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@doranngeville_radial_inreview], and set in the context of leaf phenology at these sites.

@@

For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively). 
Specifically, greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing 1-8 days per degree Celsius during the CTW (Figs. `r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).

@@

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the sustained increase in carbon sequestration that is anticipated in most models (Fig. `r schematic_summary`)[@zani_increased_2020].
Because of this, temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that these models are underestimating future overall atmospheric CO~2~.
(*see [GitHub issue 56 on representation in models](https://github.com/EcoClimLab/growth_phenology/issues/56)*)

# (Content to incorporate)

@delgado_differences_2020

From Neil: @zhang_drought_2021; @zhang_extended_2020

Geng et al. 2020 GCB- Climate warming increases spring phenological differences among temperate trees.

N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion? - - RP trees BB = 25% and 75% leaf development = 50% growth [@doranngeville_radial_inreview]. Are DP trees leafing up earlier? If yes, how much earlier? Similar to 25,50,75% DOY retreat?

@@

Leaf phenology data for the years ### - ### were extracted for SCBI and Harvard forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].

>>>>

**Title:** Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity */annual increments*

@@

DOY_timing_caption <- "**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. (**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES? (KAT thinks so)** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50). )" #needs attention!!!

DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declinde significantly with mean $T_{max}$ during the critical temperature window (CTW). For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands. Relationsips for additional growth parameters (Fig. 1a) aare given in Extended Data Figure 4. (**DOYrmax will be dropped.** [GitHub Issue 23](https://github.com/EcoClimLab/growth_phenology/issues/23))" #needs attention!!!

@@

*The introduction and summary paragraph need work, and removal of overlap. Organizational/content suggestions are welcome, but we're not at the stage of word-smithing.*

@@

In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence [REFS].
The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@zani_increased_2020].
However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].

We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview]. *(We need to make sure we're not missing anything.)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
Thus, tree-ring records do not necessarily corroborate findings that warming spring temperatures lengthen the growing season and increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America.

First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@doranngeville_radial_inreview], and set in the context of leaf phenology at these sites.

@@

For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively). 
Specifically, [greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing [1-8 days per degree Celsius](https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs. `r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).

@@

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the sustained increase in carbon sequestration that is anticipated in most models (Fig. `r schematic_summary`)[@zani_increased_2020].
Because of this, temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that these models are underestimating future overall atmospheric CO~2~.
(*see [GitHub issue 56 on representation in models](https://github.com/EcoClimLab/growth_phenology/issues/56)*)
m

@@

Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].
